### 2017-06-23 paper/compositional.tex

Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
\programbody \textbf{ od}$. We denote by $\loops(\program)$ the set of all 
loops in $\program$ there are on the top level of $\programbody$, i.e. they are 
nested loops of $\program$ nested directly "below" the outer loop. We define an 
un-nested slice of $\program$ to be a \PP{} $\slice(\program)$ obtained from 
$\program$ by removing all loops in $\loops(\program)$ and replacing them with 
skip statements. We also define the set 
$\alloops(\program)$ of all loops in $\program$, i.e. 
$\alloops(\program):=\loops(\program)\cup \{\loops(\program')\mid \program'\in 
\loops(\program)\}$. A 
formal definition of 
both functions is given in \AppendixMaterial, we illustrate them in the 
following example.

@@

\textbf{[PETR: INTRODUCE SETS OF INITIAL PROGRAM VALUATIONS RATHER THAN A 
CONCRETE VALUATION]}

Each measurable map for $\lem$ can be, in a natural way, restricted to a 
measurable map $\lem_{\program'}$ for $\program'\in \loops(\program)$, 
since each pCFG $\pCFG_{\program'}$ is a sub-pCFG of $\pCFG_{\program}$. 
Similarly, a measurable map $\lem$ for $\program$ can be restricted to a 
measurable map $\lem_{\slice}$ for $\slice(\program)$, since 
$\pCFG_{\slice(\program)}$ is produced from $\pCFG_{\program}$ essentially by 
removing certain locations (provided that we identify the locations forming 
heads of nested loops with locations preceding skip statements), and 
$\lem_{\slice}$ is defined to be equal to $\lem$ in all the locations that 
remained.

>>>>

Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
\programbody \textbf{ od}$. We denote by $\loops(\program)$ the set of all 
loops in $\program$ there are on the top level of $\programbody$, i.e. they are 
nested loops of $\program$ nested directly "below" the outer loop. We define an 
un-nested slice of $\program$ to be a \PP{} $\slice(\program)$ obtained from 
$\program$ by removing all loops in $\loops(\program)$ and replacing them with 
skip statements. We also define the set 
$\alloops(\program)$ of all loops in $\program$, i.e. 
$\alloops(\program):=\loops(\program)\cup \{\loops(\program')\mid \program'\in 
\loops(\program)\}$. The above definitions do not specify the initialization 
preamble for the sub-programs $\program'\in\alloops(\program)$: we can deem 
this pre-amble empty, i.e. consider these sub-programs with any initial 
valuation, or, provided that some invariant map $\inv$ for $\program$ is 
given, we can set the set of possible initial valuations of a sub-program 
$\program'$ to be the set $\inv(\loc^{\lin}_{\program'})$, where 
$\inv(\loc^{\lin}_{\program'})$ is the location of $\program$ corresponding to 
the head of the loop $\program'$.
A formal definition of 
both functions is given in \AppendixMaterial, we illustrate them in the 
following example.

@@

Each measurable map for $\lem$ can be, in a natural way, restricted to a 
measurable map $\lem_{\program'}$ for $\program'\in \loops(\program)$, 
since each pCFG $\pCFG_{\program'}$ is a sub-pCFG of $\pCFG_{\program}$. 
Similarly, a measurable map $\lem$ for $\program$ can be restricted to a 
measurable map $\lem_{\slice}$ for $\slice(\program)$, since 
$\pCFG_{\slice(\program)}$ is produced from $\pCFG_{\program}$ essentially by 
removing certain locations (provided that we identify the locations forming 
heads of nested loops with locations preceding skip statements), and 
$\lem_{\slice}$ is defined to be equal to $\lem$ in all the locations that 
remained. Similarly,

### 2017-06-23 paper/prelims.tex
preparing to adjust the definition of pCFG

\noindent{\em The Syntax of Probabilistic Programs {(\PP s)}.}
We consider the standard syntax for probabilistic programs,
which encompasses basic programming mechanisms such as assignment statement 
(indicated by `:='), while-loop, if-branch. Expressions appear on right-hand sides of assignments, and predicates act as loop guards and conditions in if-then-else statements. We also consider basic probabilistic mechanisms 
such as probabilistic branch (indicated by `prob') and random sampling (e.g. 
$x:=\textbf{sample(}\mathrm{Uniform}[-2,1]\textbf{)}$ assigns to $x$ a random 
number 
uniformly sampled from interval $[-2,1]$). We also allow constructs for 
(demonic) non-determinism, in particular 
non-deterministic branching indicated by `\textbf{if }$\star$ \textbf{then...}' construct and non-deterministic assignment.  
Variables (or identifiers) of a probabilistic program are of \emph{real} type, i.e., 
values of the variables are real numbers. 
We also assume that assume that each \PP{} $\program$ is preceded by an 
initialization preamble 
in 
which 
each variable appearing in $\program$ is assigned some concrete number.
the program variables are initialized.

@@

Due to space restrictions, details of syntax (such as grammar) are relegated to 
the Appendix.
For an example see Figure~\ref{fig:invariant-running}.

@@

\smallskip
\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into four 
pairwise
disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
non-deterministic branching, 
probabilistic branching, deterministic, and assignment locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinit$ is an initial 
\emph{assignment vector};
% POKRACOVAT ODSUD
\item $\transitions\subseteq\locs\times\locs$ is a transition relation;
%whose members are tuples of
%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
%variable index}, and $u$ 
%is an 
%\emph{update element}, which can 
%be one of the following mathematical objects: %\begin{itemize}
%%\item 
%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%%\item 
%(b)~a distribution $d\in \mathcal{D}$; or
%%%\item 
%(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\updates$ is a function assigning to each transition outgoing from an 
assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
\emph{target 
	variable index} and $\up$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~a Borel-measurable function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%\item 
(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$;
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

@@

\smallskip\noindent{\em Executions and reachable configurations.}
We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
configuration $(\loc,\vec{x})$ if 
%A \emph{finite path} (or
%\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
%configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
%for each
%$0 \leq i < k$ 
there is a transition  $\tau=(\loc,\loc')$ enabled in
$(\loc,\vec{x})$ 
%such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
%satisfies one of the following:
and $\vec{x}'$ satisfies the following:
\begin{compactitem}
\item if $\loc$ is not an assignment location, then $\vec{x}'=\vec{x}$;
\item if $\loc$ is an assignment location with $\updates(\tau)=(j,\up)$, then 
$\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ satisfies one of the 
following 
depending on the type of $\up$:
\begin{compactitem}
\item if $\up$ is a Borel-measurable function, then $a=\up(\vec{x})$;
\item if $\up$ is an integrable\footnote{A distribution on some numerical 
domain 
is integrable if its expected value exists and is finite. In particular, each 
Dirac distribution is integrable.} distribution $d$, then 
$a\in \support(d)$;
\item if $\up$ is a set, then $a$ is some element of $\up$.
\end{compactitem}
\end{compactitem}
A \emph{finite path} (or
\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
for each
$0 \leq i < k$ the configuration $(\loc_{i+1},\vec{x}_{i+1})$ is a successor of 
$(\loc_i,\vec{x}_i)$.
A \emph{run} (or \emph{execution}) in
$\pCFG$ is an infinite sequence of configurations whose every finite
prefix is a finite path.
A configuration $(\loc,\vec{x})$ is {\em reachable} from the initial 
configuration
$(\locinit,\vecinit)$
if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
$(\loc,\vec{x})$. We denote by $\Conf_\pCFG,\Fpath_\pCFG$ and $\Run_\pCFG$ the 
sets of all configurations, finite paths and runs in $\pCFG$, respectively, 
dropping the index $\pCFG$ when known from the context.

@@

\smallskip\noindent{\em Stochastic process.}
A pCFG $\pCFG$ together with a scheduler $\sigma$ define a stochastic 
process which produces a random run 
$(\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)(\loc_2,\vec{x}_2)\cdots$. The evolution 
of 
this process can be informally described as follows: we start in the initial 
configuration, i.e. $(\loc_{0},\vec{x}_0)=(\locinit,\vecinit)$. 
Now assume that $i$ 
steps have elapsed, i.e. a finite path 
$\genpath_i=(\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i)$ has 
already 
been produced. Then a successor configuration $(\loc_{i+1},\vec{x}_{i+1})$ is 
chosen as follows:
\begin{compactitem}
\item
If $\loc_i$ is a non-deterministic location, then 
$(\loc_{i+1},\vec{x}_{i+1})$ is sampled according to scheduler $\sigma$, i.e. 
from 
the distribution $\sigma(\genpath_i)$.
\item
If $\loc_i$ is an assignment location (but not a non-deterministic one) with, 
there 
is exactly one transition $\tau=(\loc_i,\loc')$ outgoing from it and we put 
$\loc_{i+1}=\loc'$. Denoting $\updates(\tau)=(j,\up)$, the vector 
$\vec{x}_{i+1}$ is 
then defined as 
$\vec{x}_{i+1}=\vec{x}_{i}(j\leftarrow a)$ where $a$ is chosen depending on 
$\up$:
\begin{compactitem}
\item If $u$ is a function $u\colon
\Rset^{|\pvars|}\rightarrow \Rset$, then $a=f(\vec{x}_{i})$.
\item If $\up$ is a distribution $d$, then $a$ 
is sampled from $d$.
\end{compactitem}
\item In all other cases we have $\vec{x}_{i+1}=\vec{x}_i$, and $\loc_{i+1}$ is 
determined as follows:
\begin{itemize}
	\item If $\loc_i$ is a probabilistic branching location, then a transition 
	$(\loc_i,\loc')$ is sampled from $\probdist_{\loc_i}$ and we put
	$\loc_{i+1}=\loc'$;
	\item If $\loc_i$ is deterministic, then there is exactly one transition 
	$(\loc_i,\loc')$ enabled in $(\loc_i,\vec{x}_i)$, in which case we put 
	$\loc_{i+1}=\loc'$.
\end{itemize}
\end{compactitem}

@@

The above intuitive explanation can be formalized by showing that
each pCFG $\pCFG$ together with a scheduler $\sigma$ uniquely determines a 
certain 
probabilistic 
space $(\OmegaRun,\natfilt,\probm^{\sigma})$ in which $\OmegaRun$ is a set of 
all 
runs in $\pCFG$, and a stochastic process 
$\pCFG^{\sigma}=\{\cfg{\sigma}{i}\}_{i=0}^{\infty}$ in this space 
such 
that for each run $\run\in \OmegaRun$ we have that $\cfg{\sigma}{i}(\run)$ is 
the 
$i$-th configuration on $\run$ (i.e., $\cfg{\sigma}{i}$ is a random vector 
$(\lr{\sigma}{i},\vr{\sigma}{i})$ with  $\lr{\sigma}{i}$ taking 
values in $\locs$ and $\vr{\sigma}{i}$ being a random vector of dimension 
$|\pvars|$ consisting of real-valued random variables). The sigma-algebra 
$\natfilt$ is the smallest (w.r.t. inclusion) sigma algebra under which all 
the functions $\cfg{\sigma}{i}$, for all $i\geq 0$, 
are $\natfilt$-measurable (i.e., for each $\cfg{\sigma}{i}$ and each measurable 
set of configurations $\confset\in\mathcal{F}_{\Conf}$ it holds 
$\{\run\mid\cfg{\sigma}{i}(\run)\in \confset\}\in \natfilt$). Equivalently, 
$\natfilt$ can be defined as a sigma algebra generated by all set of runs of 
the form $F\times \Conf^{\infty}$, where $F\in \mathcal{H}$ is a measurable set 
of finite paths. The 
probability 
measure 
$\probm^{\sigma}$ is such that for each $i$, the distribution of 
$\cfg{\sigma}{i}$ reflects the aforementioned way in which runs are randomly 
generated. The formal construction 
of $\natfilt$ and $\probm^{\sigma}$ proceeds via the standard \emph{cylinder 
construction}~\cite[Theorem 2.7.2]{Ash:book} and is
somewhat 
technical, hence we omit it. 
We denote by $\E^\sigma$ the expectation operator in probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma})$.

@@

\smallskip\noindent{\em Termination, reachability, and termination time.}
In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. 
We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 
$\confset$ if it  contains a configuration from $\confset$. A run 
\emph{terminates} if it reaches a 
configuration whose first component (i.e. a location of $\pCFG_P$) is the 
location
$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
executing $P$. 
To each set of configurations $\confset$ we can assign a random variable 
$\treach{\confset}$ 
such that for each run $\run$ the value $\treach{\confset}(\run)$ represents 
the first 
point in time when 
the current configuration on $\run$ is in $\confset$. If a run $\run$ does 
\emph{not} 
reach a set $\confset$, then 
$\treach{\confset}(\run)=\infty$. We call $\treach{\confset}$ the 
\emph{reachability time} of $\confset$. In particular, if $\confset$ is the set 
of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
called a \emph{termination time}, as it returns the number of steps after which 
$\run$ terminates. Since termination time is an important concept on its own, 
we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

\begin{compactenum}
\item \emph{Almost-sure termination:} A probabilistic program $P$ is 
almost-surely (a.s.) 
terminating if under each scheduler $\sigma$ it holds that 
$\probm^{\sigma}(\{\run\mid \run \text{ terminates}\}) = 1$, or equivalently, 
if for each $\sigma$ it holds $\probm^{\sigma}(\ttime<\infty)=1$. In 
almost-sure termination question for $P$ we aim to prove that $P$ 
is 
almost-surely terminating.
\item \emph{Positive termination:} A probabilistic program $P$ is 
positively
terminating if under each scheduler $\sigma$ it holds that 
$\E^{\sigma}(\{\ttime\})<  \infty$. In 
positive termination question for $P$ we aim to prove that $P$ 
is 
positively terminating. Note that each positively terminating program is also 
a.s. terminating, but the converse does not hold.
\end{compactenum}

We also define corresponding questions for the more general reachability 
concept.

\begin{compactenum}
\item \emph{Almost-sure reachability:} For a set $C$ of configurations of a
probabilistic program $P$, prove (if possible) that under each scheduler 
$\sigma$ it holds that
$\probm^{\sigma}(\treach{C}<\infty)=1$.
\item \emph{Positive reachability:} For a set $C$ of configurations of 
a probabilistic program $P$, prove (if possible) that under each scheduler 
$\sigma$ it holds that
$\E^{\sigma}(\treach{C})<\infty$.
\end{compactenum}

Since termination is a special case of reachability, each solution to the 
almost-sure or probabilistic reachability questions provides solution for the 
corresponding termination questions.

>>>>

\noindent{\em The Syntax of Probabilistic Programs {(\PP s)}.}
We consider the standard syntax for probabilistic programs,
which encompasses basic programming mechanisms such as assignment statement 
(indicated by `:='), while-loop, if-branch. Expressions appear on right-hand sides of assignments, and predicates act as loop guards and conditions in if-then-else statements. We also consider basic probabilistic mechanisms 
such as probabilistic branch (indicated by `prob') and random sampling (e.g. 
$x:=\textbf{sample(}\mathrm{Uniform}[-2,1]\textbf{)}$ assigns to $x$ a random 
number 
uniformly sampled from interval $[-2,1]$). We also allow constructs for 
(demonic) non-determinism, in particular 
non-deterministic branching indicated by `\textbf{if }$\star$ \textbf{then...}' construct and non-deterministic assignment.  
Variables (or identifiers) of a probabilistic program are of \emph{real} type, i.e., 
values of the variables are real numbers. 
We also assume that assume that each \PP{} $\program$ is preceded by a preamble 
specifying possible initial values of program variables: the preamble consists 
of a single predicate characterizing possible initial valuations.

@@

Due to space restrictions, details of syntax (such as grammar) are relegated to 
\AppendixMaterial.
For an example see Figure~\ref{fig:invariant-running}.

@@

\smallskip
\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinitset,\transitions,\updates,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into four 
pairwise
disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
non-deterministic branching, 
probabilistic branching, deterministic, and assignment locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinitset$ is a set of initial 
\emph{assignment vectors};
% POKRACOVAT ODSUD
\item $\transitions\subseteq\locs\times\locs$ is a transition relation;
%whose members are tuples of
%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
%variable index}, and $u$ 
%is an 
%\emph{update element}, which can 
%be one of the following mathematical objects: %\begin{itemize}
%%\item 
%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%%\item 
%(b)~a distribution $d\in \mathcal{D}$; or
%%%\item 
%(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\updates$ is a function assigning to each transition outgoing from an 
assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
\emph{target 
	variable index} and $\up$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~a Borel-measurable function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%\item 
(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$;
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

@@

\smallskip\noindent{\em Executions and reachable configurations.}
We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
configuration $(\loc,\vec{x})$ if 
%A \emph{finite path} (or
%\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
%configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
%for each
%$0 \leq i < k$ 
there is a transition  $\tau=(\loc,\loc')$ enabled in
$(\loc,\vec{x})$ 
%such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
%satisfies one of the following:
and $\vec{x}'$ satisfies the following:
\begin{compactitem}
\item if $\loc$ is not an assignment location, then $\vec{x}'=\vec{x}$;
\item if $\loc$ is an assignment location with $\updates(\tau)=(j,\up)$, then 
$\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ satisfies one of the 
following 
depending on the type of $\up$:
\begin{compactitem}
\item if $\up$ is a Borel-measurable function, then $a=\up(\vec{x})$;
\item if $\up$ is an integrable\footnote{A distribution on some numerical 
domain 
is integrable if its expected value exists and is finite. In particular, each 
Dirac distribution is integrable.} distribution $d$, then 
$a\in \support(d)$;
\item if $\up$ is a set, then $a$ is some element of $\up$.
\end{compactitem}
\end{compactitem}
A \emph{finite path} (or
\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
$\loc_0=\locinit$, $\vec{x}_0\in\vecinitset$, and 
%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
for each
$0 \leq i < k$ the configuration $(\loc_{i+1},\vec{x}_{i+1})$ is a successor of 
$(\loc_i,\vec{x}_i)$.
A \emph{run} (or \emph{execution}) in
$\pCFG$ is an infinite sequence of configurations whose every finite
prefix is a finite path.
A configuration $(\loc,\vec{x})$ is {\em reachable} from the initial 
configuration
$(\locinit,\vecinit)$ (where, $\vecinit\in \vecinitset$)
if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
$(\loc,\vec{x})$. We denote by $\Conf_\pCFG,\Fpath_\pCFG$ and $\Run_\pCFG$ the 
sets of all configurations, finite paths and runs in $\pCFG$, respectively, 
dropping the index $\pCFG$ when known from the context.

@@

\smallskip\noindent{\em Stochastic process.}
A pCFG $\pCFG$ together with a scheduler $\sigma$ and initial valuation 
$\vecinit\in\vecinitset$ define a stochastic 
process which produces a random run 
$(\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)(\loc_2,\vec{x}_2)\cdots$. The evolution 
of 
this process can be informally described as follows: we start in the initial 
configuration, i.e. $(\loc_{0},\vec{x}_0)=(\locinit,\vecinit)$. 
Now assume that $i$ 
steps have elapsed, i.e. a finite path 
$\genpath_i=(\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i)$ has 
already 
been produced. Then a successor configuration $(\loc_{i+1},\vec{x}_{i+1})$ is 
chosen as follows:
\begin{compactitem}
\item
If $\loc_i$ is a non-deterministic location, then 
$(\loc_{i+1},\vec{x}_{i+1})$ is sampled according to scheduler $\sigma$, i.e. 
from 
the distribution $\sigma(\genpath_i)$.
\item
If $\loc_i$ is an assignment location (but not a non-deterministic one) with, 
there 
is exactly one transition $\tau=(\loc_i,\loc')$ outgoing from it and we put 
$\loc_{i+1}=\loc'$. Denoting $\updates(\tau)=(j,\up)$, the vector 
$\vec{x}_{i+1}$ is 
then defined as 
$\vec{x}_{i+1}=\vec{x}_{i}(j\leftarrow a)$ where $a$ is chosen depending on 
$\up$:
\begin{compactitem}
\item If $u$ is a function $u\colon
\Rset^{|\pvars|}\rightarrow \Rset$, then $a=f(\vec{x}_{i})$.
\item If $\up$ is a distribution $d$, then $a$ 
is sampled from $d$.
\end{compactitem}
\item In all other cases we have $\vec{x}_{i+1}=\vec{x}_i$, and $\loc_{i+1}$ is 
determined as follows:
\begin{itemize}
	\item If $\loc_i$ is a probabilistic branching location, then a transition 
	$(\loc_i,\loc')$ is sampled from $\probdist_{\loc_i}$ and we put
	$\loc_{i+1}=\loc'$;
	\item If $\loc_i$ is deterministic, then there is exactly one transition 
	$(\loc_i,\loc')$ enabled in $(\loc_i,\vec{x}_i)$, in which case we put 
	$\loc_{i+1}=\loc'$.
\end{itemize}
\end{compactitem}

@@

The above intuitive explanation can be formalized by showing that
each pCFG $\pCFG$ together with a scheduler $\sigma$ and initial valuation 
$\vecinit\in \vecinitset$ uniquely determine a 
certain 
probabilistic 
space $(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ in which $\OmegaRun$ 
is a set of 
all 
runs in $\pCFG$, and a stochastic process 
$\pCFG^{\sigma}=\{\cfg{\sigma}{i}\}_{i=0}^{\infty}$ in this space 
such 
that for each run $\run\in \OmegaRun$ we have that $\cfg{\sigma}{i}(\run)$ is 
the 
$i$-th configuration on $\run$ (i.e., $\cfg{\sigma}{i}$ is a random vector 
$(\lr{\sigma}{i},\vr{\sigma}{i})$ with  $\lr{\sigma}{i}$ taking 
values in $\locs$ and $\vr{\sigma}{i}$ being a random vector of dimension 
$|\pvars|$ consisting of real-valued random variables). The sigma-algebra 
$\natfilt$ is the smallest (w.r.t. inclusion) sigma algebra under which all 
the functions $\cfg{\sigma}{i}$, for all $i\geq 0$, 
are $\natfilt$-measurable (i.e., for each $\cfg{\sigma}{i}$ and each measurable 
set of configurations $\confset\in\mathcal{F}_{\Conf}$ it holds 
$\{\run\mid\cfg{\sigma}{i}(\run)\in \confset\}\in \natfilt$). Equivalently, 
$\natfilt$ can be defined as a sigma algebra generated by all set of runs of 
the form $F\times \Conf^{\infty}$, where $F\in \mathcal{H}$ is a measurable set 
of finite paths. The 
probability 
measure 
$\probm^{\sigma}_{\vecinit}$ is such that for each $i$, the distribution of 
$\cfg{\sigma}{i}$ reflects the aforementioned way in which runs are randomly 
generated. The formal construction 
of $\natfilt$ and $\probm^{\sigma}_{\vecinit}$ proceeds via the standard 
\emph{cylinder 
construction}~\cite[Theorem 2.7.2]{Ash:book} and is
somewhat 
technical, hence we omit it. 
We denote by $\E^\sigma_{\vecinit}$ the expectation operator in probability 
space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$.

@@

\smallskip\noindent{\em Termination and termination time.}
In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. 
We say that a run 
\emph{terminates} if it reaches a 
configuration whose first component (i.e. a location of $\pCFG_P$) is the 
location
$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
executing $P$. 
We define a random variable $\ttime$
such that for each run $\run$ the value $\ttime(\run)$ represents 
the first 
point in time when 
the current location is $\loc_P^{\lout}$. If a run $\run$ does 
\emph{not} 
terminate, then 
$\ttime(\run)=\infty$. We call $\ttime$ the 
\emph{termination time} of $\program$. 
%In particular, if $\confset$ is the set 
%of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
%terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
%called a \emph{termination time}, as it returns the number of steps after 
%which 
%$\run$ terminates. Since termination time is an important concept on its own, 
%we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

\begin{compactenum}
\item \emph{Almost-sure termination:} A probabilistic program $P$ is 
almost-surely (a.s.) 
terminating if under each scheduler $\sigma$ and for each initial valuation 
$\vecinit\in\vecinitset$ it holds that 
$\probm^{\sigma}_{\vecinit}(\{\run\mid \run \text{ terminates}\}) = 1$, or 
equivalently, 
if for each $\sigma$ it holds $\probm^{\sigma}(\ttime<\infty)=1$. In 
almost-sure termination question for $P$ we aim to prove that $P$ 
is 
almost-surely terminating.
\item \emph{Positive termination:} A probabilistic program $P$ is 
positively
terminating if under each scheduler $\sigma$ and for each initial valuation 
$\vecinit\in\vecinitset$ it holds that 
$\E^{\sigma}_{\vecinit}(\{\ttime\})<  \infty$. In 
positive termination question for $P$ we aim to prove that $P$ 
is 
positively terminating. Note that each positively terminating program is also 
a.s. terminating, but the converse does not hold.
\end{compactenum}

%We also define corresponding questions for the more general reachability 
%concept. 
%\begin{compactenum}
%\item \emph{Almost-sure reachability:} For a set $C$ of configurations of a
%probabilistic program $P$, prove (if possible) that under each scheduler 
%$\sigma$ it holds that
%$\probm^{\sigma}(\treach{C}<\infty)=1$.
%\item \emph{Positive reachability:} For a set $C$ of configurations of 
%a probabilistic program $P$, prove (if possible) that under each scheduler 
%$\sigma$ it holds that
%$\E^{\sigma}(\treach{C})<\infty$.
%\end{compactenum}
%Since termination is a special case of reachability, each solution to the 
%almost-sure or probabilistic reachability questions provides solution for the 
%corresponding termination questions.
